---
title: "Lab 5"
author: "Joshua Palter"
output: pdf_document
---


We will work with the diamonds dataset from last lecture:

```{r}
pacman::p_load(ggplot2) #this loads the diamonds data set too
?diamonds
diamonds$cut =      factor(diamonds$cut, ordered = FALSE)
diamonds$color =    factor(diamonds$color, ordered = FALSE)
diamonds$clarity =  factor(diamonds$clarity, ordered = FALSE)
skimr::skim(diamonds)
```

Given the information above, what are the number of columns in the raw X matrix?

9

Verify this using code:

```{r}
ncol(diamonds)
```

Would it make sense to use polynomial expansions for the variables cut, color and clarity? Why or why not?

no because it's not continuous

Would it make sense to use log transformations for the variables cut, color and clarity? Why or why not?

no because it's not continuous

In order to ensure there is no time trend in the data, randomize the order of the diamond observations in D:.

```{r}
diamonds = diamonds[sample(nrow(diamonds)),]
```

Let's also concentrate only on diamonds with <= 2 carats to avoid the issue we saw with the maximum. So subset the dataset. Create a variable n equal to the number of remaining rows as this will be useful for later. Then plot it.

```{r}
diamonds = diamonds[diamonds$carat <= 2,]
n = nrow(diamonds)
ggplot(diamonds, aes(x = carat, y = price)) + geom_point()
```

Create a linear model of price ~ carat and gauge its in-sample performance using s_e.

```{r}
mod1 = lm(price ~ carat,diamonds)
summary(mod1)$sigma
```

Create a model of price ~ clarity and gauge its in-sample performance

```{r}
mod = lm(price ~ clarity,diamonds)
summary(mod)$sigma
```

Why is the model price ~ carat substantially more accurate than price ~ clarity?

because it's both continuous? and more correlated

Create a new transformed feature ln_carat and plot it vs price.

```{r}
diamonds$ln_carat = log(diamonds$carat)
ggplot(diamonds, aes(x = ln_carat, y = price)) + geom_point()
```

Would price ~ ln_carat be a better fitting model than price ~ carat? Why or why not?

No, it has a higher standard deviation

Verify this by comparing R^2 and RMSE of the two models:

```{r}
mod1 = lm(price ~ carat, diamonds)
summary(mod1)$sigma
mod2 = lm(price~ln_carat,diamonds)
summary(mod2)$sigma
```

Create a new transformed feature ln_price and plot its estimated density:


```{r}
diamonds$ln_price = log(diamonds$price)
ggplot(diamonds) + geom_histogram(aes(x = ln_price), binwidth = 0.01)
```


Now plot it vs carat.

```{r}
ggplot(diamonds, aes(x = carat, y = ln_price)) + 
  geom_point()
```

Would ln_price ~ carat be a better fitting model than price ~ carat? Why or why not?

Yes, it has a lower standard deviation

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}
mod3 = lm(ln_price ~ carat, diamonds)
y_hat = exp(mod3$fitted.values)
sse = sum((y_hat - diamonds$price)^2)
rmse = sqrt(sse/(n - 2))
summary(mod1)$sigma
summary(mod2)$sigma
summary(mod3)$sigma
```

We just compared in-sample statistics to draw a conclusion on which model has better performance. But in-sample statistics can lie! Why is what we did valid?

We are only using one feature!! We aren't overfitting with one feature 52k datapoints (probably)

Plot ln_price vs ln_carat.

```{r}
ggplot(diamonds, aes(x = ln_carat, y = ln_price)) + 
  geom_point()
```

Would ln_price ~ ln_carat be the best fitting model than the previous three we considered? Why or why not?

Yeah it looks more like a line :)
(It has lower standard deviation)

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}
mod4 = lm(ln_price ~ ln_carat, diamonds)
y_hat = exp(mod4$fitted.values)
sse = sum((y_hat - diamonds$price)^2)
sqrt(sse/(n-2))
summary(mod3)$sigma
summary(mod4)$sigma
```

Compute b, the OLS slope coefficients for this new model of ln_price ~ ln_carat.

```{r}
coef(mod4)
#Model A
```

Interpret b_1, the estimated slope of ln_carat.

% change in price is around % change in x

Interpret b_0, the estimated intercept.

The predicted price of a "ghost diamond", which has no weight (The minimum limit of a logical function)

Create other features ln_x, ln_y, ln_z, ln_depth, ln_table.

```{r}
diamonds$ln_x = log(diamonds$x)
diamonds$ln_y = log(diamonds$y)
diamonds$ln_z = log(diamonds$z)
diamonds$ln_depth = log(diamonds$depth)
diamonds$ln_table = log(diamonds$table)
```

From now on, we will be modeling ln_price (not raw price) as the prediction target. 

Create a model (B) of ln_price on ln_carat interacted with clarity and compare its performance with the model (A) ln_price ~ ln_carat.

```{r}
ggplot(diamonds, aes(x = ln_carat, y = ln_price, color = clarity)) + geom_point()
moda = lm(ln_price ~ ln_carat, diamonds)
summary(moda)$sigma
#Model B
modb = lm(ln_price ~ ln_carat * clarity, diamonds)
summary(modb)$sigma
```

Which model does better? Why?

Model B as it adds another non-orthogonal dimension to colspace[x]/another feature

Create a model of (C) ln_price on ln_carat interacted with every categorical feature (clarity, cut and color) and compare its performance with model (B)

```{r}
modc = lm(ln_price ~ ln_carat * (clarity + cut + color), diamonds)
summary(modc)$sigma
#Model C
```

Which model does better? Why?

Model C as we're adding another non-orthogonal dimension into the colspace[X]/another feature

Create a model (D) of ln_price on every continuous feature (logs of carat, x, y, z, depth, table) interacted with every categorical feature (clarity, cut and color) and compare its performance with model (C).

```{r}
#Model D
diamonds = diamonds[diamonds$x > 0 & diamonds$z > 0, ]
diamonds[is.infinite(diamonds$ln_x),]
modd = lm(ln_price ~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table) * (clarity + cut + color), diamonds)
summary(modd)$sigma
```

Which model does better? Why?

Model D as we're adding another non-orthogonal dimension into the colspace[X]/another feature

What is the p of this model D? Compute with code.

```{r}
modd$rank
ncol(model.matrix(~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table) * (clarity + cut + color), diamonds))
```

Create model (E) which is the same as before except create include the raw features interacted with the categorical features and gauge the performance against (D).

```{r}
#Model E
mod_e = lm(ln_price ~ (carat + x + y + z + depth + table) * (clarity + cut + color), diamonds)
summary(mod_e)$sigma
```

Which model does better? Why?

Model E as we're adding another non-orthogonal dimension into the colspace[X]/another feature

Create model (F) which is the same as before except also include also third degree polynomials of the continuous features interacted with the categorical features and gauge performance against (E). By this time you're getting good with R's formula syntax!

```{r}
#Model F
modf = lm(ln_price ~ (poly(carat,3) + poly(x,3) + poly(y,3) + poly(z,3) + poly(depth,3) + poly(table,3)) * (clarity + cut + color), diamonds)
summary(modf)$sigma
```

Which model does better? Why?

Model F as we're adding another non-orthogonal dimension into the colspace[X]/another feature

Can you think of any other way to expand the candidate set curlyH? Discuss.

Allow the functions allowed to be within cubic functions to be within quadratic functions.

We should probably assess oos performance now. Sample 2,000 diamonds and use these to create a training set of 1,800 random diamonds and a test set of 200 random diamonds. Define K and do this splitting:

```{r}
K = 10 #1/10 of 2,000 is 2000
nsamp = 2000
set.seed(89)
D = diamonds[sample(1:nrow(diamonds), nsamp),]
Dtrain = D[1:((1-1/K) * nsamp),]
Dtest = D[((1-1/K) * nsamp + 1) : nsamp, ]
```

Compute in and out of sample performance for models A-F. Use s_e as the metric (standard error of the residuals). Create a list with keys A, B, ..., F to store these metrics. Remember the performances here will be worse than before since before you're using nearly 52,000 diamonds to build a model and now it's only 1,800! 

```{r}
inSampleRMSE = list()
outOfSampleRMSE = list()
moda = lm(ln_price ~ ln_carat, diamonds)
inSampleRMSE[['A']] = summary(moda)$sigma

modb = lm(ln_price ~ ln_carat * clarity, diamonds)
inSampleRMSE[['B']] = summary(modb)$sigma


modc = lm(ln_price ~ ln_carat * (clarity + cut + color), diamonds)
inSampleRMSE[['C']] = summary(modc)$sigma


modd = lm(ln_price ~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table) * (clarity + cut + color), diamonds)
inSampleRMSE[['D']] = summary(modd)$sigma

mod_e = lm(ln_price ~ (carat + x + y + z + depth + table) * (clarity + cut + color), diamonds)
inSampleRMSE[['E']] = summary(mod_e)$sigma

modf = lm(ln_price ~ (poly(carat,3) + poly(x,3) + poly(y,3) + poly(z,3) + poly(depth,3) + poly(table,3)) * (clarity + cut + color), diamonds)
inSampleRMSE[['F']] = summary(modf)$sigma


outOfSampleRMSE[['A']] = sd(Dtest$ln_price ~ predict(moda,Dtest))
outOfSampleRMSE[['B']] = sd(Dtest$ln_price ~ predict(modb,Dtest))
outOfSampleRMSE[['C']] = sd(Dtest$ln_price ~ predict(modc,Dtest))
outOfSampleRMSE[['D']] = sd(Dtest$ln_price ~ predict(modd,Dtest))
outOfSampleRMSE[['E']] = sd(Dtest$ln_price ~ predict(mod_e,Dtest))
outOfSampleRMSE[['F']] = sd(Dtest$ln_price ~ predict(modf,Dtest))


cbind(
  unlist(insampleRMSE),
  unlist(oosRMSE)
)
#TO-DO
```

You computed oos metrics only on n_* = 200 diamonds. What problem(s) do you expect in these oos metrics?

They are likely going to be less accurate than what the actual prediction is, given there's a smaller sample size for training

To do the K-fold cross validation we need to get the splits right and crossing is hard. I've developed code for this already. Run this code.

```{r}
temp = rnorm(n)
folds_vec = cut(temp, breaks = quantile(temp, seq(0, 1, length.out = K + 1)), include.lowest = TRUE, labels = FALSE)
head(folds_vec, 200)
```

Comment on what it does and how to use it to do a K-fold CV:

It basically breaks the training dataset up into 10 pieces and shuffles around which of those ten pieces are going to be the testing dataset. To use it in K-Fold CV, we can take which number it outputs and use it to test our dataset.

Do the K-fold cross validation for model F and compute the overall s_e and s_s_e. 

```{r}
#TO-DO
```

Does K-fold CV help reduce variance in the oos s_e? Discuss.

#TO-DO

Imagine using the entire rest of the dataset besides the 2,000 training observations divvied up into slices of 200. Measure the oos error for each slice on Model F in a vector s_e_s_F and compute the s_s_e_F and also plot it.

```{r}
#TO-DO
ggplot(data.frame(s_e_s_F = s_e_s_F)) + geom_histogram(aes(x = s_e_s_F))
```